---
format: pdf

fontsize: 11pt
geometry: margin=1in
header-includes:
- \usepackage{fancyhdr}
- \pagenumbering{gobble}
- \usepackage{hyperref}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}

execute: 
  echo: false
  warning: false
  error: false

params:
  YEAR: 2024
---

```{r}
#| label: Package Install and Data Import
#| output: false

source("../data/Import_Packages_Data.R")
```

\colorbox{black}{\rlap{\huge\bfseries\textcolor{white}{U.S. Fish \& Wildlife Service}}\hspace{\linewidth}\hspace{-2\fboxsep}}\vspace{-.15cm}
\begin{flushleft}
\large\textbf{Lower Great Lakes Fish and Wildlife Conservation Office}\vspace{.15cm}
\newline
\large\textit{Aquatic Invasive Species Program}\vspace{.15cm}
\newline
\large\textbf{Early Detection Monitoring Metabarcoding Surviellance}\vspace{.25cm}

\large\textbf{Program Objective}

Aquatic invasive species (AIS) pose a serious threat to the Great Lakes. Prevention, early detection, and rapid response provide the most effective and potentially successful means to minimize harm to the environment and costs associated with stopping the spread of nonnative species. The U.S. Fish and Wildlife Service (USFWS) developed the Strategic Framework for the Early Detection of Non-native Fishes in the Great Lakes identifying early detection as a priority action. The objective of the Early Detection and Monitoring (EDM) program is to detect novel invasive species and document AIS range expansions while populations are low and geographically isolated - providing management actions a better opportunity of success. 

\large\textbf{Metabarcoding Implementation}

Along with traditional fisheries surveys (electrofishing and netting), molecular surveillance using environmental DNA (eDNA) has proven to be a valuable tool in AIS monitoring. Metabarcoding, a multi-taxa eDNA method, now allows for fish community level surveillance across a broad range of species, both native, non-native, and AIS watchlist. The USFWS will be implementing a annual metabarcoding AIS surveillance program that will monitor both broad-spectrum AIS watchlist species and targeted species-specific AIS of interest. Objectives for both surveillance purposes are defined as follows:

\begin{itemize}
\item\textbf{Broad-Spectrum:} Monitoring will focus on AIS watchlist species across high priority locations (HPLs) comprised of harbors, rivers and embayments of the lower Great Lakes as determined by a harbors, rivers, and embayments of the Great Lakes as determined by a \href{https://cfpub.epa.gov/si/si_public_record_Report.cfm?dirEntryId=350136&Lab=CCTE}{risk-based            prioritization framework for Aquatic Invasive Species (AIS) in the Great Lakes basin} (Figure 1). Highest priority HPLs will     get the largest sample efforts, with sample allocations trickling down across the other lower ranked priority sites (Table 1). 
\item\textbf{Species-Specific:} Monitoring will also focus on particular AIS species while still employing a multi-taxa approach (Figure 2; Table 2). 
\begin{itemize}
\item\textbf{Tench} (\textit{Tinca tinca}), an Eurasian cyprinid, will have continued eDNA monitoring along the invasion front in the Massena, NY region of St. Lawrence River. Massena, NY maintains a High priority status due to the range expansion of Tench and binational partner interest in monitoring that invasion front. 
\item\textbf{Grass Carp} (\textit{Ctenopharyngodon idella}), an Asian cyprinid, will have continued and expanded effort across areas within Lake Ontario and Eastern Lake Erie. Sampling areas were selected based off high priority designation and locations that had habitat conditions preferred by grass carp.
\end{itemize}
\end{itemize}

\large\textbf{Recapping 2024 Metabarcoding Efforts}

Partnerships and objectives from last field season to inform fture partners of past goals and partner involvement.

\end{flushleft}

\newpage

\large\textbf{Broad-Spectrum AIS Surviellance}
\vspace{1.5cm}

```{r}
#| label: EDM Priority Location Map
#| output: asis
#| eval: true
#| fig-height: 7.5
#| fig-width: 9.5
#| fig-cap: "Map of broad-spectrum AIS monitoring locations across the lower Great Lakes with their corresponding priority level (Red: High, Orange: Moderate, Blue: Low)."
 
 ## Create Sampling Locations

points <- metaB %>%
  filter(Purpose == "Early Detection Monitoring") %>% 
  select(Location, Priority, x, y)

sites <- st_as_sf(points, coords = c("x", "y"), crs = 4326, agr = "constant")

## Plot location map
 
Priority_map <- ggplot() + 
  geom_sf(data = cropped, size = .5, color = "grey60") +
  geom_sf(data = NY_sf, size = .5, color = "grey60", fill = NA) +
  #geom_sf(data = US_sf, size = .5, color = "grey60", fill = NA) +
  #geom_sf(data = sites, size = 3, shape = 21, fill = "red3", color = "red1", stroke = 1.25, alpha = .5) +
  geom_point(data = points, aes(x = x, y = y, color = Priority), size = 3, stroke = 1.25, alpha = .65) +
  scale_color_manual(values=c("#fa0a0a","#f5ad37", "#2bd2e9")) +
  geom_text_repel(data = points, aes(x = x, y = y, label = Location), size = 3.2, point.padding = 1, max.overlaps = Inf) + 
  annotate(geom = "text", x = -78.1, y = 43.65, label = "Lake Ontario", 
    fontface = "italic", color = "grey60", size = 3) +
  annotate(geom = "text", x = -79.59, y = 42.65, label = "Lake Erie", 
    fontface = "italic", color = "grey60", size = 2.5) +
  coord_sf()+
  #geom_sf_text(data = sites, aes(label = Location), color = "black", size = 2.6, nudge_y = -.15)  +
  annotation_scale(location = "bl", style = "ticks", line_col = "black", text_col = "black") +
  theme(
        panel.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x=element_text(colour="black"),
        axis.text.y=element_text(colour="black"),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.line = element_line(colour = "black"),
        panel.border = element_rect(colour = "black", fill=NA, size=.5)
  )

Priority_map
```

\newpage
\blandscape

```{r}
#| label: EDM Table
#| output: asis
#| eval: true

# Pack Rows 
basin_rows <- metaB %>% 
  filter(Purpose == "Early Detection Monitoring") %>%
  group_by(Basin, Location) %>% 
  summarise() %>% 
  ungroup() %>% 
  count(Basin, name = "Rows")

# Generate table

EDM_tbl <- metaB %>% 
  filter(Purpose == "Early Detection Monitoring") %>% 
  rename("Priority*" = Priority, "FWS Only" = Metabarcoding_fws, "FWS & Partners" = Metabarcoding_all) %>% 
  arrange(Basin, `Priority*`) %>% 
  select(-Basin, -x, -y, -Purpose, -Target_Species, -Partnerships) %>% 
  #select(2, 1, 3, 4, 5, 6) %>% 
  adorn_totals(,,,,-`Fish Rank`) %>% 
  kable(., format = "latex", digits = 2, booktabs = TRUE, longtable = FALSE, caption = "Priority broad-spectrum AIS monitoring locations across the lower Great Lakes.", align = 'lcccc') %>%
  kable_styling(latex_options = "hold_position") %>%
  add_header_above(c(" " = 1, "Ranking out of 388 Locations" = 2, "eDNA Sample Capacity Scenarios" = 2)) %>%
  kable_styling(latex_options = c("repeat_header")) %>% 
  kable_styling(full_width = FALSE) %>%
  kable_styling(font_size = 12) %>%
  #column_spec(1, bold=T) %>% 
  #kable_styling(latex_options = "striped") %>%
  #collapse_rows(1, latex_hline = "major", valign = "top") %>% 
  pack_rows(index=c("Lake Erie" = pluck(basin_rows, 2, 1), "Lake Ontario" = pluck(basin_rows, 2, 2))) %>%
  footnote(symbol  = "Priority designation was assigned based on rankings from the risk-based prioritization framework.", threeparttable = T) 

EDM_tbl
```
\elandscape
\newpage

\large\textbf{Species-Specfifc Targeted AIS Surviellance}
\vspace{1.5cm}

```{r}
#| label: Target Priority Location Map
#| output: asis
#| eval: true
#| fig-height: 7.5
#| fig-width: 9.5
#| fig-cap: "Map of targeted AIS species specific locations across the lower Great Lakes with their corresponding priority level (Red: High, Orange: Moderate, Blue: Low)."
 
 ## Create Sampling Locations

points <- metaB %>%
  filter(Purpose == "Targeted") %>% 
  select(Location, Priority, x, y)

sites <- st_as_sf(points, coords = c("x", "y"), crs = 4326, agr = "constant")

## Plot location map
 
Target_map <- ggplot() + 
  geom_sf(data = cropped, size = .5, color = "grey60") +
  geom_sf(data = NY_sf, size = .5, color = "grey60", fill = NA) +
  #geom_sf(data = US_sf, size = .5, color = "grey60", fill = NA) +
  #geom_sf(data = sites, size = 3, shape = 21, fill = "red3", color = "red1", stroke = 1.25, alpha = .5) +
  geom_point(data = points, aes(x = x, y = y, color = Priority), size = 3, stroke = 1.25, alpha = .65) +
  scale_color_manual(values=c("#fa0a0a","#f5ad37", "#2bd2e9")) +
  geom_text_repel(data = points, aes(x = x, y = y, label = Location), size = 3.2, point.padding = 1, max.overlaps = Inf) + 
  annotate(geom = "text", x = -78.1, y = 43.65, label = "Lake Ontario", 
    fontface = "italic", color = "grey60", size = 3) +
  annotate(geom = "text", x = -79.59, y = 42.65, label = "Lake Erie", 
    fontface = "italic", color = "grey60", size = 2.5) +
  coord_sf()+
  #geom_sf_text(data = sites, aes(label = Location), color = "black", size = 2.6, nudge_y = -.15)  +
  annotation_scale(location = "bl", style = "ticks", line_col = "black", text_col = "black") +
  theme(
        panel.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x=element_text(colour="black"),
        axis.text.y=element_text(colour="black"),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.line = element_line(colour = "black"),
        panel.border = element_rect(colour = "black", fill=NA, size=.5)
  )

Target_map
```

\newpage
\blandscape

```{r}
#| label: Targeted Table
#| output: asis
#| eval: true

# Pack Rows 
basin_rows <- metaB %>% 
  filter(Purpose == "Targeted") %>%
  group_by(Basin, Location) %>% 
  summarise() %>% 
  ungroup() %>% 
  count(Basin, name = "Rows")

# Generate table

Target_tbl <- metaB %>% 
  filter(Purpose == "Targeted") %>% 
  rename("Priority*" = Priority, "FWS Only" = Metabarcoding_fws, "FWS & Partners" = Metabarcoding_all, "Target Species" = Target_Species) %>% 
  arrange(Basin, `Priority*`) %>% 
  select(-Basin, -x, -y, -Purpose, -`Fish Rank`) %>% 
  select(1,2,3,4,6,5) %>% 
  adorn_totals() %>% 
  kable(., format = "latex", digits = 2, booktabs = TRUE, longtable = FALSE, caption = "Priority targeted species-specific AIS monitoring locations across the lower Great Lakes.", align = 'lcccll') %>%
  kable_styling(latex_options = "hold_position") %>%
  add_header_above(c(" " = 2, "eDNA Sample Capacity Scenarios" = 2, " " = 2)) %>%
  kable_styling(latex_options = c("repeat_header")) %>% 
  kable_styling(full_width = FALSE) %>%
  kable_styling(font_size = 12) %>%
  #column_spec(1, bold=T) %>% 
  #kable_styling(latex_options = "striped") %>%
  collapse_rows(5, latex_hline = "major", valign = "top") %>% 
  pack_rows(index=c("Lake Erie" = pluck(basin_rows, 2, 1), "Lake Ontario" = pluck(basin_rows, 2, 2))) %>%
  footnote(symbol  = "Priority designation was assigned based on rankings from the risk-based prioritization framework.", threeparttable = T) 

Target_tbl
```

\elandscape
