---
format: pdf

fontsize: 11pt
geometry: margin=0.75in
header-includes:
- \usepackage{fancyhdr}
- \pagenumbering{gobble}
- \usepackage{helvet}
- \usepackage{tcolorbox}
- \usepackage{pdflscape}
- \usepackage{float}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}

execute: 
  echo: false
  warning: false
  error: false
  cache: false
---

```{r}
#| label: Package Install and Data Import
#| output: false
#| cache: true

source(here::here('data', 'Import_Packages_Data.R'))
```

\colorbox{black}{\rlap{\huge\bfseries\textcolor{white}{U.S. Fish \& Wildlife Service}}\hspace{\linewidth}\hspace{-2\fboxsep}}\vspace{-.15cm}
\begin{flushleft}
\large\textbf{Lower Great Lakes Fish and Wildlife Conservation Office}\vspace{.15cm}
\newline
\large\textit{Aquatic Invasive Species Program}\vspace{.15cm}
\newline
\large\textbf{Early Detection Monitoring Metabarcoding Surviellance}\vspace{.25cm}

\large\textbf{Program Objective}

Aquatic invasive species (AIS) pose a serious threat to the Great Lakes. Prevention, early detection, and rapid response provide the most effective and potentially successful means to minimize harm to the environment and costs associated with stopping the spread of nonnative species. The U.S. Fish and Wildlife Service (USFWS) developed the Strategic Framework for the Early Detection of Non-native Fishes in the Great Lakes identifying early detection as a priority action. The objective of the Early Detection and Monitoring (EDM) program is to detect novel invasive species and document AIS range expansions while populations are low and geographically isolated - providing management actions a better opportunity of success. 

**All yours Colleen!**

\newpage

\large\textbf{Broad-Spectrum AIS Surviellance}
\vspace{1.5cm}

```{r}
#| label: All Priority Location Map
#| output: asis
#| eval: true
#| fig-height: 7.5
#| fig-width: 9.5
#| fig-cap: "Map of broad-spectrum AIS monitoring locations across the lower Great Lakes with their corresponding priority level (Red: High, Orange: Moderate, Blue: Low)."
 
 ## Create Sampling Locations

points <- metaB %>%
  filter(Purpose == "Early Detection Monitoring") %>% 
  select(Location, Priority, x, y)

sites <- st_as_sf(points, coords = c("x", "y"), crs = 4326, agr = "constant")

## Plot location map
 
Priority_map <- ggplot() + 
  geom_sf(data = cropped, size = .5, color = "grey60") +
  geom_sf(data = NY_sf, size = .5, color = "grey60", fill = NA) +
  #geom_sf(data = US_sf, size = .5, color = "grey60", fill = NA) +
  #geom_sf(data = sites, size = 3, shape = 21, fill = "red3", color = "red1", stroke = 1.25, alpha = .5) +
  geom_point(data = points, aes(x = x, y = y, color = Priority), size = 3, stroke = 1.25, alpha = .65) +
  scale_color_manual(values=c("#fa0a0a","#f5ad37", "#2bd2e9")) +
  geom_text_repel(data = points, aes(x = x, y = y, label = Location), size = 3.2, point.padding = 1, max.overlaps = Inf) + 
  annotate(geom = "text", x = -78.1, y = 43.65, label = "Lake Ontario", 
    fontface = "italic", color = "grey60", size = 3) +
  annotate(geom = "text", x = -79.59, y = 42.65, label = "Lake Erie", 
    fontface = "italic", color = "grey60", size = 2.5) +
  coord_sf()+
  #geom_sf_text(data = sites, aes(label = Location), color = "black", size = 2.6, nudge_y = -.15)  +
  annotation_scale(location = "bl", style = "ticks", line_col = "black", text_col = "black") +
  theme(
        panel.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x=element_text(colour="black"),
        axis.text.y=element_text(colour="black"),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.line = element_line(colour = "black"),
        panel.border = element_rect(colour = "black", fill=NA, size=.5)
  )

Priority_map
```

\newpage
\blandscape

```{r}
#| label: All metaB sampling
#| output: asis
#| eval: true

# Pack Rows 
basin_rows <- metaB %>% 
  filter(Purpose == "Early Detection Monitoring") %>%
  group_by(Basin, Location) %>% 
  summarise() %>% 
  ungroup() %>% 
  count(Basin, name = "Rows")

# Generate table

EDM_tbl <- metaB %>% 
  filter(Purpose == "Early Detection Monitoring") %>% 
  rename("Priority*" = Priority, "FWS Only" = Metabarcoding_fws, "FWS & Partners" = Metabarcoding_all) %>% 
  arrange(Basin, `Priority*`) %>% 
  select(-Basin, -x, -y, -Purpose, -Target_Species, -Partnerships) %>% 
  #select(2, 1, 3, 4, 5, 6) %>% 
  adorn_totals(,,,,-`Fish Rank`) %>% 
  kable(., format = "latex", digits = 2, booktabs = TRUE, longtable = FALSE, caption = "Priority broad-spectrum AIS monitoring locations across the lower Great Lakes.", align = 'lcccc') %>%
  kable_styling(latex_options = "hold_position") %>%
  add_header_above(c(" " = 1, "Ranking out of 388 Locations" = 2, "eDNA Sample Capacity Scenarios" = 2)) %>%
  kable_styling(latex_options = c("repeat_header")) %>% 
  kable_styling(full_width = FALSE) %>%
  kable_styling(font_size = 12) %>%
  #column_spec(1, bold=T) %>% 
  #kable_styling(latex_options = "striped") %>%
  #collapse_rows(1, latex_hline = "major", valign = "top") %>% 
  pack_rows(index=c("Lake Erie" = pluck(basin_rows, 2, 1), "Lake Ontario" = pluck(basin_rows, 2, 2))) %>%
  footnote(symbol  = "Priority designation was assigned based on rankings from the risk-based prioritization framework.", threeparttable = T) 

EDM_tbl
```
\elandscape
\newpage